Penalized linear mixed models for structured genetic data

### 2021-03-03 main.tex
5.2 case study figure 6 results

%------------------------------%
\subsection{\anna{Genome-wide analysis}}
%------------------------------%

To illustrate the statistical properties demonstrated by our simulations on a more realistic scale, we present an analysis of the full Empirical genomic data ($n=588, p=526,266$) from \citet{larkin2015objectives} with a synthetic phenotype as a pseudo-real data case study. We generate the synthetic phenotype as in \eqref{eqn:sim} using 50,000 causal SNPs, a 1:1 SNR ($\eta = 0.5$), and substantial environmental effects ($\xi = 0.8$).

%------------------------------%
\subsection{\anna{Software}}
%------------------------------%
\label{sec:software}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results} \label{sec:results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%------------------------------%
\subsection{PC-lasso and LMM-lasso increase accuracy}
\label{sec:sim-mse}
%------------------------------%

@@

In addition to MSE, these methods' performance was also assessed in terms of variable selection using the true positive rate (TPR) for the first 50 variables each method allowed into the model. Throughout, we use ``true positive'' to mean both that the SNP was selected and that its direction was estimated correctly. Table \ref{tab:var_sel} displays the average TPR for the same settings as in Figure \ref{fig:mse}. The TPRs echo the MSE results in that lasso and LMM-lasso perform comparably in the absence of environmental confounding across all data types, with PC-lasso slightly less accurate at identifying causal SNPs. In the presence of environmental confounding, both LMM-lasso and PC-lasso achieve substantial improvements in TPR over the ordinary lasso. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Empirical data sets than in the 1D Linear Admixture data.

@@

%------------------------------%
\subsection{PC-lasso and LMM-lasso reduce selection bias}
%------------------------------%

@@

%------------------------------%
\subsection{\anna{LMM-lasso outperforms lasso when analyzing genome-wide data}}
%------------------------------%

\begin{figure}[H]
\centering
\includegraphics[scale = 0.9]{figures/genomewide_prc.png}
\includegraphics[scale = 0.9]{figures/genomewide_prc2.png}
\caption{Precision values for each method and number of selected SNPs from the Empirical genome-wide analysis. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
\label{fig:gw_prc}
\end{figure}
\anna{should the y-axes on this and the below fig match? i.e. both be number of selected snps, or both be lambda indices?}\\
\anna{Try number of snps selected for x-axis, keep y-axis as precision}\\
\anna{Also why is pc-lasso doing so terribly? \textbf{depends on how you're defining precision... selects a lot of true snps but often they have the wrong sign}}\\

\begin{figure}[H]
\centering
\includegraphics[scale = 0.9]{figures/genomewide_pc1_corr.png}
\caption{Pathwise absolute correlation of selected, non-causal SNPs with the first principal component (PC1) of the data in Empirical genome-wide analysis. Plotted line is the mean absolute correlation of all incorrectly selected SNPs for a particular $\lambda$ value, with shaded region indicating SE bounds. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
\label{fig:gw_pc1_corr}
\end{figure}
\anna{Should y-axis here be \# of selected snps to match above fig?}

@@

@@

@@

\anna{\textbf{I don't think we should bring up the idea of CV/subsampling on original vs transformed data in this paper at all - I think it would just make things more confusing without a lot of additional explanation.}}

>>>>

%------------------------------%
\subsection{\anna{Software}}
%------------------------------%
\label{sec:software}
\anna{Should this be here as a methods subsection or should it get its own section after discussion, before data availability?}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{\anna{Simulation results}} \label{sec:results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%------------------------------%
\subsection{PC-lasso and LMM-lasso increase estimation accuracy}
\label{sec:sim-mse}
%------------------------------%

@@

In addition to MSE, these methods' performance was also assessed in terms of variable selection using the true positive rate (TPR) for the first 50 variables each method allowed into the model. \anna{Throughout, we use ``true positive'' to mean both that the SNP was selected and that its direction was estimated correctly.} Table \ref{tab:var_sel} displays the average TPR for the same settings as in Figure \ref{fig:mse}. The TPRs echo the MSE results in that lasso and LMM-lasso perform comparably in the absence of environmental confounding across all data types, with PC-lasso slightly less accurate at identifying causal SNPs. In the presence of environmental confounding, both LMM-lasso and PC-lasso achieve substantial improvements in TPR over the ordinary lasso. As with the MSE data, these improvements are more substantial in the Independent Subpopulations and Empirical data sets than in the 1D Linear Admixture data.

@@

%------------------------------%
\subsection{PC-lasso and LMM-lasso reduce selection bias} \label{sec:sel_bias}
%------------------------------%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{\anna{Genome-wide case study}} \label{sec:gw_cs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To illustrate the statistical properties demonstrated by our simulations on a more realistic genome-wide scale, we present an analysis of the full Empirical genomic data ($n=588, p=526,266$) from \citet{larkin2015objectives} and a synthetic phenotype as a pseudo-real data case study. We generate the synthetic phenotype as in \eqref{eqn:sim} using 50,000 causal SNPs, a 1:1 SNR ($\eta = 0.5$), and substantial environmental effects ($\xi = 0.8$). In order to compare all methods' performance fairly, we present results for the entirety of each method's $\lambda$ path.

%------------------------------%
\subsection{\anna{LMM-lasso identifies causal SNPs with greater precision than lasso}}
%------------------------------%
\begin{figure}[H]
\centering
\includegraphics[scale = 0.9]{figures/genomewide_prc.png}
\caption{Precision values for each method and number of selected SNPs from the Empirical genome-wide analysis. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
\label{fig:gw_prc}
\end{figure}

$$\text{Precision }= 1 - \frac{\#\{ \beta = 0 \cap \widehat{\beta} \ne 0  \}}{\#\{ \widehat{\beta} \ne 0 \}} = \frac{\#\{ \beta \ne 0 \cap \widehat{\beta} \ne 0  \}}{\#\{ \widehat{\beta} \ne 0 \}}$$

\begin{figure}[H]
\centering
\includegraphics[scale = 0.9]{figures/genomewide_prc2.png}
\caption{Precision values for each method and number of selected SNPs from the Empirical genome-wide analysis. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
\label{fig:gw_prc2}
\end{figure}

$$\text{Precision }= \frac{\#\{ \beta > 0 \cap \widehat{\beta} \ne 0  \}}{\#\{ \widehat{\beta} \ne 0 \}}$$

\anna{Why is pc-lasso doing so terribly? \textbf{original bad seed + depends on how you're defining precision... selects a lot of true snps but often they have the wrong sign}}\\

\noindent\anna{should the y-axes on this and the below fig match? i.e. both be number of selected snps, or both be lambda indices?}\\

%------------------------------%
\subsection{\anna{PC-lasso and LMM-lasso reduce selection bias in genome-wide data}}
%------------------------------%

A central issue explored in this work has been the potential for population stratification and environmental confounding to lead to the identification of spurious associations. In Section~\ref{sec:sel_bias} we characterized the three lasso methods (ordinary, PC and LMM) in terms of their susceptibility to this type of selection bias in a simplified setting of two subpopulations. We now seek to examine these methods' performance in a genome-wide, pathwise manner.

Figure \ref{fig:gw_pc1_corr} displays the mean absolute correlation of the non-causal SNPs selected for a particular $\lambda$ value with the first principal component (PC1) of the Empirical data for each method across the full $\lambda$ path. The corresponding shaded regions indicate the SE about the mean; overall these bounds are largest early on in each $\lambda$ path, where only a few SNPs are be selected, but narrow as $\lambda$ increases and more variables are allowed to enter the model.\anna{Do I need to explain why we are using correlation with the first PC as a proxy for MAF difference or is this obvious?} The absolute correlation with PC1 is used as a measure of how population-informative the selected SNPs are given the complex Empirical data structure compared to that in Section~\ref{sec:sel_bias}.

\begin{figure}[H]
\centering
\includegraphics[scale = 0.9]{figures/genomewide_pc1_corr.png}
\caption{Pathwise absolute correlation of selected, non-causal SNPs with the first principal component (PC1) of the data in Empirical genome-wide analysis. Plotted line is the mean absolute correlation of all incorrectly selected SNPs for a particular $\lambda$ value, with shaded region indicating SE bounds. Single replication case study with $\eta = 0.5, \xi = 0.8$.}
\label{fig:gw_pc1_corr}
\end{figure}

The ordinary lasso is clearly most susceptible to selecting non-causal SNPs due to their association with population structure. This corresponds to the simulation results shown in Section~\ref{sec:sel_bias}, though in Figure \ref{fig:gw_pc1_corr} we more plainly see that this bias persists across the $\lambda$ path. Likewise, PC-lasso most effectively controls this type of selection bias. This is to be expected given that we are assessing the population-informative nature of incorrectly selected SNPs via correlation with PC1, which is explicitly adjusted for in the PC-lasso method. LMM-lasso yields a substantial improvement compared to the ordinary lasso in controlling this selection bias. It should be noted that the Empirical data's RRM in Figure~\ref{fig:empirical} shows that its structure is predominantly coarse. Given our results in Section~\ref{sec:sim-fine-coarse}, we might expect a still more substantial improvement over the ordinary lasso in the analysis of more finely structured genome-wide data.

@@

\anna{Add discussion of genome-wide results}

@@

\textbf{\anna{Add reference to that paper that mentions this challenge explicitly}}

@@
